---
title: "TITLE"
author: "Amaya McMillin, Ashley Fong, Constantine Golematis, Rayne Lejano"
format: html
editor: visual
embed-resources: true
code-tools: true
code-fold: true
---

# 1 Project Proposal & Data

**Data:**

```{r, include = FALSE}
library(tidyverse)
library(gt)

seafood <- read_csv("fisfod_cons_pc.csv")
lex <- read_csv("lex.csv")
```

In this analysis, we are exploring two datasets: **life expectancy at birth** and **fish and seafood consumption per capita**. Could eating seafood be correlated with longevity?

Data Source:

We have sourced both of our datasets from **Gapminder** which is an independent foundation that works to combine data from multiple sources and fill in gaps in data in order to create more complete datasets that provide us with a chronological story of information. In both these datasets, Gapminder defines the geogrphic boundaries of the country variable as where the border currently lay.

Life Expectancy Dataset:

The life expectancy dataset provides data on the **number of years** a newborn infant is expected to live based on the **year of birth** and \*\*country of birth\*. This data ranges from 1800 to 2100 and contains information from 196 different countries.

Fish and Seafood Consumption Dataset:

The fish and seafood consumption dataset provides data on the **average supply of fish and seafood across the population** and is measured in **kilograms per person per year** based on the **year** and **country**. In this dataset, the supply is defined as the amount available for human consumption in that country including food imported and produced, but excluding food exported, used for animals, or distribution wasted (not household waste). Based on this definition of consumption, it is likely that there are overestimates of actual consumption. This data ranges from 1961 to 2018 and contains information from 170 different countries.

**Hypothesis:**

We hypothesize that higher seafood consumption is associated with a greater life expectancy across countries. Seafood is linked to various health benefits, including improved cognitive function and improved cardiovascular health. By examining the relationship between the two variables, we can explore if this hypothesis reigns true.

## 1.1 Data Cleaning

-   how is missing data handled
-   how are variables defined
-   observations dropped (how many and why)

Many values in both data sets are recorded as N/A. To ensure a complete analysis, we have chosen to replace these missing values with the average of their respective rows.

```{r}
lex[-1] <- lex[-1] |>
  mutate(across(everything(), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

seafood[-1] <- seafood[-1] |>
  mutate(across(everything(), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))
```

## 1.2 Pivoting Longer

In order to join the data sets, we must first pivot the data sets into a long format. Each data set contains a single column for the country, while every other column represents a different year. If we attempted to join the data sets in this wide format, we would end up with multiple columns for each year, making it difficult to determine which data set each value belongs to.

```{r}
life_expectancy_long <- lex |>
  pivot_longer(cols = -country,
               names_to = 'year',
               values_to = 'life_expectancy') |>
  mutate(year = as.factor(year))

seafood_long <- seafood |>
  pivot_longer(cols = -country, 
               names_to = 'year',
               values_to = 'kg_seafood') |>
  mutate(year = as.factor(year))


```

## 1.3 Joining Data

Now that we have pivoted the data, we can join the new data sets to perform our analysis. We will be using an inner join so we can keep only the data that is present in both data sets. The life expectancy data set contains data for years 1800 to 2100, while the seafood data set contains data on years 1961-2018. An inner join would ensure we work with data that overlaps in both data sets.

```{r}
full_data <- life_expectancy_long |>
  inner_join(seafood_long, by = c('country', 'year'))
```
